Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-02 analysis/paper/paper.Rmd
complete draft

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia rivers have created favorable conditions for fishes known to consume juvenile salmonids. The native Northern Pikeminnow are one of these fishes and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating throu the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine predation by Northern Pikeminnow on Chinook Salmon in a slow-water reach of the Salmon River, Idaho known as Deadwater Slough. The reach was (at least partially) formed by the failure of a small mining dam in a nearby drainage during the late 19th century and contains historically important overwinter rearing habitat is a migration corridor for Chinook Salmon and other Endangered Species Act-listed species. We estimate the abundance of Northern Pikeminnow during peak emigrations to be ***insert numbers here***. Using a bioenergetics approach we estimate the consumption potential of the Northern Pikeminnow population to be ***key results here***. This resulted in the potential consumption of ***numbers*** Chinook Salmon emigrants. Using a sensitivity analysis, we evaluated the potential impact to adult returns of local Chinook Salmon populations to be between ***number with CIs***. It is likely that predation of juvenile Chinook Salmon in Deadwater Slough has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River major population group, thereby hindering local recovery efforts.

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created favorable conditions for Northern Pikeminnow. We estimate that more than ***value*** Northern Pikeminnow occupy Deadwater Slough, resulting in a remarkable density given the size of Deadwater Slough. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon. We estimate that predation from Northern Pikeminnow in Deadwater Slough may reduce adult returns by as much as 146 or 26%, and therefore, has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor’s Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from Caitlin Alcott and Inter-Fluve. This manuscript benefitted from a review by Sean Gibbs. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor’s Office of Species Conservation.

>>>>

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindinger population recovery efforts in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia river have created favorable conditions for fishes known to consume juvenile salmonids. The native Northern Pikeminnow are one of these fished and have been estimate to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The reach was (at least partially) formed by the failure of a small mining dam in a nearby drainage during the late 19th century and contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon and other Endangered Species Act-listed species including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 27,874 in the fall and 19,499 in the spring during the peak emigrations of juvenile Chinook Salmon. Assuming a scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is Chinook Salmon, we estimated the Northern Pikeminnow population to consume 95,120 juvenile Chinook Salmon, annually. In the most conservative scenario where 30% of diet is fish and 30% of fish are juvenile Chinook Salmon 47,079 juveniles are consumed. Using a sensitivity analysis, we further considered potential impacts to adult returns to local populations. Given the size of the Salmon River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable and predation from the population likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group, thereby hindering local recovery efforts.

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created favorable conditions for Northern Pikeminnow. Even under the most conservative scenarios, we estimate more than 10,000 Northern Pikeminnow occupy Deadwater Slough, resulting in a remarkable density given the size of Deadwater Slough. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon. We estimate that, at a minimum, Northern Pikeminnow in Deadwater Slough consume greater than 24,000 juvenile Chinook salmon (the lower 95% CI of the most conservative scenario evaluated). It is more likely that this value is closer to 100,000, annually, creating a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor’s Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from Caitlin Alcott and Inter-Fluve. This manuscript benefited from a review by Sean Gibbs. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor’s Office of Species Conservation.
